---
title: "case_study_RT_accuracy"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, error=TRUE, cache = FALSE)
```

```{r libraries, include=FALSE}
library(tidyverse)
library(magrittr)
library(lme4)
```

## Experiment design

Describe the design of the experiment in your own words:
By employing a brief exposure paradigm, this study looks at adaptation to difficult-to-understand speech by manipulating the form in each condition (speech in noise vs foreign-accented speech) and measuring accuracy and RT performance on the task.

```{r load-data}
d.all <- read.csv(file = "~/BMB_JHS/LSCI253M/data/xie_data_full.csv") %>%
  select(PartOfExp, Trial, Filename, Word, Response, RT, WorkerId, Condition, List, Speaker, VisualProbeType, Error, CorrectResponse, Block, BaselineRT, ListOrder, ListID, Phase, subjMean_BlockRT, BaselineRT_raw) %>%
   filter(Condition %in% 
           c("Mandarin-accented English (same)",
             "Control (native accent)" 
         ))
```

## Data cleaning

### 1. Examine RT distribution

Examine the distribution of RT (subjMean_BlockRT) across subjects. Does it make sense?

It reflects the data and some extreme participant mean RTs. For example, participant "7c22ded74c03d60a95d8b0cd59e2a3cc" has a mean RT of 62092.667 for the practice block (that's plotted in the histogram) and also another mean RT of 16956.333 for the main block (also observed in the plot). I would say that because it includes outliers from the data, it makes sense.

```{r distribution-of-subj-wise-mean-RTs-before-exclusions, fig.cap="Distribution of subjects mean RTs by Block and Condition, prior to outlier exclusions.\\label{fig:distribution-of-subj-wise-mean-RTs-before-exclusions}"}
rt_dist <- d.all %>% 
  select(WorkerId, PartOfExp, Block, Condition, Trial, Error, RT, subjMean_BlockRT, BaselineRT, BaselineRT_raw) # why are we piping and what? include RTs here, too? 
  
# rt_dist

hist(rt_dist$subjMean_BlockRT, xlab = "Reaction Times", main= "Distribution of mean RTs with outliers included")

```

## 2. Data exclusion

Describe the procedure you take to exclude outliers (subjects, trials, etc.).

### Exclusion by subject
Describe your exclusion criteria based on a subject's performance.

e.g., We want to identify and remove subjects who consistently registered slow response times because they did not perform the task faithfully (e.g., multi-tasking) or because their computer equipment did not provide reliable recording of RTs over the web. 


```{r outlier-exclusion-subject}
## ----------------------------------------
# identify *eligible* subjects 

# how many RT-based subject exclusions in total --> 0

# Create a boxplot
boxplot(rt_dist$BaselineRT)

# Identify the outliers by looking at the baseline RTs. Those with a slower baseline RT 
outliers <- boxplot(rt_dist$BaselineRT, plot = FALSE)$out

# Add labels to the outliers so we know which to exclude
text(rep(1, length(outliers)), outliers, labels = outliers, pos = 4)

# Exclude 6 specific participants that came up as outliers in the box plot
#participants_to_exclude <- c("adc7f84030b38b867c3ed0f120581ba1", "094b7ceb77cd85e63ae18c67790b0ddd", "9f9e46f1bf14e98550be2abbc6213c9f", "1a80f32d763fab9b525dbe7782d2fe29", "a130e2284329819b8f5c696bee7015fc", "7511c8bdb426c2f475ebd2383ae17014")
filtered_data <- subset(data, !participant %in% participants_to_exclude)

# View the resulting filtered data
#filtered_data

#BEFORE EXCLUDING - let's look at box plot statistics to see if these really are outliers

# Calculate box plot statistics
stats <- boxplot.stats(rt_dist$BaselineRT)

# Create a table of statistics
statistics_table <- data.frame(
  Min = stats$stats[1],
  Lower_Quartile = stats$stats[2],
  Median = stats$stats[3],
  Upper_Quartile = stats$stats[4],
  Max = stats$stats[5],
  Outliers = ifelse(length(stats$outliers) > 0, stats$outliers, NA))

# Display the table
print(statistics_table)
#NO OUTLIERS... no need to exclude participants then. 



# how many RT-based subject exclusions per Condition --> 0 
# Calculate statistics and outliers by condition
stats <- aggregate(subjMean_BlockRT ~ Condition, rt_dist, FUN = function(x) {
  outliers <- boxplot.stats(x)$out
  c(Mean = mean(x), Median = median(x), SD = sd(x), Outliers = ifelse(length(outliers) > 0, outliers, NA))
})

# View the resulting statistics table with outliers
stats
#                         Condition subjMean_BlockRT.Mean subjMean_BlockRT.Median subjMean_BlockRT.SD subjMean_BlockRT.Outliers
#1          Control (native accent)              1339.168                1107.000             685.344                  3020.000
#2 Mandarin-accented English (same)              1678.620                1273.250            3637.288                  3322.667

          
```

Re-examine RT distribution after subject exclusion.

```{r RT-distribution-after-outlier-removal-step1, fig.cap="...\\label{fig:RT-distribution-after-outlier-removal-step1}" }

#rt_dist %+%  # not sure what to do with this...

hist(rt_dist$subjMean_BlockRT, xlab = "Reaction Times", main= "Distribution of mean RTs without subject exclusions")


```

### Exclusion by trial with extreme RTs

The second step of outlier removal was to exclude trials with atypical RTs. Describe your exclusion criteria by trial and do a second round of exclusion.

(1) We will exclude trials with RTs that are less than 200ms or greater than 8000ms for each trial.

(2) Then, we will exlude trials with RTs that are greater than 3 standard deviations from the participant's mean RT. (Xie et al., 2018)

Q: Did trial-wise outlier exclusion disproportionately affect any experimental Conditions?

In step 1: 
13 participants from Mandarin-accented condition removed.
12 participants from Control condition removed.
In step 2:
42 participants from Mandarin-accented condition removed.
47 participants from Control condition removed.

Overall, I'd say they weren't disproportionately affected by outlier exclusion. 

```{r outlier-removal-step2, echo = FALSE}

# Filter out trials with reaction times less than 200ms & those greater than 8000ms
dat_out1 <- subset(rt_dist, RT >= 200 & RT <= 8000)
filtered_out_data1 <- subset(rt_dist, !(RT >= 200 & RT <= 8000))

# View the filtered out data
#filtered_out_data
# View the resulting filtered data
# dat_out1


# Filtering out using SD criteria
participant_stats <- dat_out1 %>%
  group_by(WorkerId) %>%
  summarise(mean_RT = mean(RT), sd_RT = sd(RT))

# Merge into dataset
dat_out1 <- dat_out1 %>%
  left_join(participant_stats, by = "WorkerId")

# Calculate participant-wise mean and standard deviation
threshold_multiplier <- 3

dat_out1 <- dat_out1 %>%
  mutate(upper_threshold = mean_RT + threshold_multiplier * sd_RT)


# Calculate the upper threshold for excluding extreme values
# Filter out participants whose values exceed the upper threshold
dat_out2 <- dat_out1 %>%
  filter(RT <= upper_threshold)

filtered_out_data2 <- dat_out1 %>%
  filter(RT > upper_threshold)

```

Q: Examine the mean RTs by block. Do they vary a lot before and after trial exclusion? Describe the effects.


## Examine RTs and Accuracy during practice and baseline (after exclusion steps 1 and 2)

Now that we've excluded extreme subject and trial outliers, we can look at the practice and baseline data to assess our high-level predictions about how participants should perform on this web-based task.

1. **One data pattern that we expect to find is that performance (both RTs and accuracy) in the practice and baseline blocks is comparable across experimental conditions.** We expect this because these blocks of the experiment were identical across conditions (i.e., native-accented stimuli presented in the clear).
    
    + ... *if performance in the **practice block** differs substantially across conditions*, we would need to consider whether the subjects in each condition were sampled from the same underlying population (e.g., did we run all conditions at approximately the sme time of day?).

    + ... *if performance in the **baseline block** differs substantially across conditions*, we would need to consider whether exposure to different types of speech during the main block of the experiment induced overall differences in task performance (in which case the baseline block doesn't provide a reliable condition-independent "baseline" for normalization purposes).

2. **A second data pattern that we expect to find is evidence of improvement (adaptation) over the course of the task.** One way this would manifest is faster RTs and increased accuracy in the post-experiment baseline block, relative to the practice phase. 


## Summary of exclusion criteria:\label{sec:summary-of-exclusion-criteria}

- Participant-level exclusions:
    + Identify participants, who across conditions, demonstrate substantially different performance in practice blocks. 
    + Identify participants, who across conditions, demonstrate substantially different performance in baseline blocks. 
    
    
- Trial-level exclusions:
    + RTs and error rates greater than 3SD from the mean.

We applied the same exclusion criteria across all RT and error analyses.

## Normalize experimental RTs relative to baseline

Now that we've completed all trial-wise RT exclusions, we can calculate _normalized_ RTs that take into account each subject's baseline speed on this task. For this procedure, we adjust the RTs on each trial by subtracting out the corresponding subject's mean RT during the baseline phase. We refer to the resulting measure as _adjusted RTs_.

```{r, echo = TRUE}
# calculate each subject's mean Baseline RT
# and subtract that value from experimental RTs
dat_out2 %<>%
  group_by(WorkerId) %>%
  mutate(
    # calculate subject-wise mean RTs during baseline block
    meanBaselineRT = mean(RT[PartOfExp == "baseline"]),
    
    # calculate normalized RTs
    AdjustedRT = RT - meanBaselineRT,
    
    # calculate subject-wise mean Adjusted RT across Blocks 1-4
    meanAdjustedRT = mean(AdjustedRT[PartOfExp == "main"])
  )
```

Now we want to check the distribution of adjuted RTs to make sure it seems reasonable, given our expectations about task performance.

Note that we expect baseline RTs to be faster on average than RTs during the experimental block, regardless of exposure condition. We expect this for two reasons. First, the baseline task occurred at the end of the experiment, after participants had adapted to the task. Second, _all_ participants heard native accented speech during the baseline phase; hence, there was no need for accent adaptation during this phase.


# Modeling strategy

## Model building and assessment
RTs were analyzed using linear mixed effects regression, as implemented in the lme4 package (version 1.1-10: Bates, Maechler, Bolker, \\& Walker, 2014) in R (R Core Team, 2014). Response accuracy (incorrect vs. correct response) was analyzed using mixed effects logistic regression (see Jaeger, 2008). All mixed effects models were specified with the maximal random effects structure justified by the experimental design: that is, by-subject and by-item random intercepts, by-subject random slopes for all design variables manipulated within subjects, and by-item random slopes for all design variables manipulated within items. If the definitionally maximal model failed to converge within ten thousand iterations, the model was systematically simplified in a step-wise fashion until the model converged. These steps involved removing correlations among random effects; dropping the random effects term with the least variance; and removing fixed effects that were inconsequential for the theory being tested (i.e., counterbalancing nuisance variables).

## Variable coding
Unless otherwise specified, all numeric predictors were centered and categorical predictors were coded as sum contrasts, in order to reduce collinearity among predictors. 

```{r prep-lmer}

# change to dat_out3 to implement 3rd outlier step
dat <- dat_out2 %>%
  filter(PartOfExp == "main") %>%
  droplevels(.)

## ------------------------------------------ 
## Define contrast coding for analyses
## ------------------------------------------ 

dat <- within(dat %>%
                mutate(Block = factor(Block)), {
  # helmert coding for Block for C&G-style analysis
  contrasts(Block) <- contr.helmert(4)
})

## ------------------------------------------ 
## EXPERIMENT 1
exp1 <- dat %>%
  within(., {
  # sum coding for accent condition
  Condition <- factor(Condition)
	contrasts(Condition) <- cbind("Accented" = c(1,-1))
	
	 # sum contrast code List (counterbalancing nuissance factor)
	List <- factor(List)
  contrasts(List) <- contr.sum(nlevels(List))
  colnames(contrasts(List)) <- rownames(contrasts(List))[1:7]
  
  # sum code ListID
  ListID <- factor(ListID)
  contrasts(ListID) <- contr.sum(nlevels(ListID))

  #sum code ListOrder
  ListOrder <- factor(ListOrder)
  contrasts(ListOrder) <- contr.sum(nlevels(ListOrder))
})
```


# Experiment 1: Adaptation to Mandarin-accented English
## Participants

Examine the number of participants per condition.

```{r examine-number-of-participants}

```
## Exp1 Response Times

Visualize the changes of RTs across blocks by condition.

```{r exp1-RTs-by-condition, fig.width = 11, fig.height = 5, fig.cap="Average RTs by exposure condition in Experiment 1.\\label{fig:exp1-RTs-by-condition}"}


# Plot the changes of RTs across blocks by condition
ggplot(dat_out2, aes(x = Block, y = RT, color = Condition, group = WorkerId)) +
  geom_line() +
  geom_point() +
  labs(x = "Block", y = "Reaction Time (ms)", color = "Condition") +
  theme_minimal()

ggplot(dat_out2, aes(x = Condition, y = RT, fill = factor(Block))) +
  geom_boxplot() +
  labs(x = "Condition", y = "Reaction Time (ms)", fill = "Block") +
  theme_minimal()


```

We assess the effect of exposure condition (Mandarin-accented English vs. control) on processing speed separately for RTs during the exposure phase and the test phase. To assess the _change_ in RTs during the course of exposure, we split the 18-trial exposure phase into three blocks of 6 trials and use the resulting Block variable as a categorical predictor of RTs. We use linear mixed-effects models to simultaneously model subject and item random effects.


#### Exposure
A linear mixed effects model was fit to adjusted RTs for correct responses during the exposure phase. 

Describe your fixed effects and random effects. Describe how each variable is coded.
  +   Fixed effects: BaselineRTs
  +   Random effects: Block and Condition
```{r exp1-byBlock-exposureRT, echo = TRUE}
# Model specification:
# by-block analysis of RTs during EXPOSURE

head(dat_out2, n=20)
```


#### Test

```{r exp1-byBlock-testRT, echo = TRUE}
# Model specification:
mod = lmer(RT~ BaselineRT + (Block|Condition), data = dat_out2)
# by-block analysis of RTs during TEST 



```